---
title: "Lecture 1: Exercise Problems and Solutions"
format: 
  html:
    theme: 
      - cosmo
    embed-resources: true
    toc: true
    number-sections: true
    execute:
      eval: false
      message: false
      warning: false
---


# Exercise: Vector

<!-- start panel:Ex1   --> 
:::{.panel-tabset .nav-pills} 
## {{< fa person-chalkboard >}} Instructions 

The following code randomly samples 30 numbers from a uniform distribution between 0 and 1, and stores the result in `x`. 

```{r}
# Run this code to work on the exercise problems.
set.seed(3746)
x <- runif(n = 30, min = 0, max = 1)
x # see what's inside x
```


**Questions**

1. Extract the 10th and the 15th elements of `x`.
2. Extract elements larger than $0.5$.
3. Replace the 10th and the 15th elements of `x` to 0.
4. If an element of `x` is larger than $0.9$, replace it with $1$.
5. Count the elements larger than $0.6$. 


## {{< fa lightbulb >}} Solutions

```{r}
# === Part 1 === #
x[c(10, 15)]

# === Part 2 === #
x[x > 0.5]

# === Part 3 === #
x[c(10, 15)] <- 0

# === Part 4 === #
x[x > 0.9] <- 1

# === Part 5 === #
sum(x > 0.6)
```

<!-- end panel:Ex1  --> 
:::

<br>

# Exercise: Matrix

<!-- start panel:matrix   --> 
:::{.panel-tabset .nav-pills} 
## Instruction## {{< fa person-chalkboard >}} Instructions 

Use the following matrix:

```{r}
set.seed(3746)
num <- runif(n = 30, min = 0, max = 1)
mat <- matrix(data = num, nrow = 6)
colnames(mat) <- c("A", "B", "C", "D", "E")
rownames(mat) <- c("a", "b", "c", "d", "e", "f")
mat # see what's inside mat
```


1. Extract the element in the 2nd row and 3rd column.
2. Extract the 2nd row.
3. Subset the rows where column "A" is larger than 0.5. (Use logical indexing).


## {{< fa lightbulb >}} Solutions
 
```{r}
# === Part 1 === #
mat[2, 3]

# === Part 2 === #
mat[2, ]

# === Part 3 === #
mat[mat[, "A"] > 0.5, ]
```
<!-- end panel:matrix  --> 
:::

<br>

# Exercise: Data Frame

<!-- start panel: ex data.frame   --> 
:::{.panel-tabset .nav-pills}  
## {{< fa person-chalkboard >}} Instructions  


We will use the built-in dataset `mtcars` for this exercise. Run the following code to load the data.

```{r}
# --- Load data --- #
data(mtcars)
?mtcars # to see the description of the yield_data

# --- Take a look at the data --- #
# head() function shows the first several rows of the data
head(mtcars)
```


1. Extract the rows corresponding to the cars with the row numbers 1, 5, and 10 using numeric indexing

2. Add a new column to the `mtcars` data frame called `power_to_weight_ratio`, which is calculated as the ratio of horsepower (`hp`) to weight (`wt`).

3. Create a new data frame called `efficient_cars` that contains cars with `mpg` greater than `20` and power-to-weight ratio less than 5.

4. (Optional) Sort the efficient_cars data frame by the `power_to_weight_ratio` column in ascending order and display the result. [Hints: (1)use `order()` function to sort the data frame. (2) Use `order(efficient_cars$power_to_weight_ratio)` as an index vector.]


## {{< fa lightbulb >}} Solutions

```{r}
# === Part 1 === #
mtcars[c(1, 5, 10), ]

# === Part 2 === #
mtcars$power_to_weight_ratio <- mtcars$hp / mtcars$wt

# === Part 3 === #
efficient_cars <- mtcars[mtcars$mpg > 20 & mtcars$power_to_weight_ratio < 5, ]

# === Part 4 === #
efficient_cars[order(efficient_cars$power_to_weight_ratio), ]
```

<!-- end panel: ex data.frame  --> 
:::

<br>

# Exercise: Vector, Comprehensive

<!-- start panel: ex --> 
:::{.panel-tabset .nav-pills}  
## {{< fa person-chalkboard >}} Instructions  

1. Create a sequence of numbers from 20 to 50 and name it `x`. Let's change the numbers that are multiples of 3 to 0.

2. `sample()` is commonly used in Monte Carlo simulation in econometrics. Run the following code to create `r`. What does it do? Use `?sample` to find out what the function does.

```{r}
#| autorun: true
set.seed(12345) #don't worry about this
r <- sample(1:100, size=20, replace = TRUE)
```

3. Find the value of mean and SD of vector `r` without using `mean()` and `sd()`
4. Figure out which position contains the maximum value of vector `r`. (use `which()` function. Run `?which()` to find out what the function does.)
5. Extract the values of `r` that are larger than 50.
6. Extract the values of `r` that are larger than 40 and smaller than 60.
7. Extract the values of `r` that are smaller than 20 or larger than 70.

## {{< fa lightbulb >}} Solutions

```{r}
# === Part 1 === #
x <- 20:50
# using `:` operator is the most basic way to create a sequence of numbers, but it only works with integer numbers with a step of 1.
# seq() function is more flexible. For example, you can create a sequence of numbers, , incremented by 0.5.
# x <- seq(from = 20, to = 50, by = 0.5)
x[x %% 3 == 0] <- 0

# === Part 2 === #
# In this code, sample() function creates a random sample of numbers with size 20 (size=20) from a range 1 to 100 (x = 1:100) allowing replacement (replace = TRUE).

# === Part 3 === #
# mean
mean_r <- sum(r) / length(r)
# SD
sd_r <- sqrt(sum((r - mean_r)^2) / (length(r) - 1))

# === Part 4 === #
max_index <- which(r == max(r))

# === Part 5 === #
r_50 <- r[r > 50]

# === Part 6 === #
r_40_60 <- r[r > 40 & r < 60]

# === Part 7 === #
r_20_70 <- r[r < 20 | r > 70]
```

<!-- end panel: ex  --> 
:::


<br>


# Exercise: Data Frame, Comprehensive

<!-- start panel: ex --> 
:::{.panel-tabset .nav-pills}  
## {{< fa person-chalkboard >}} Instructions  

1. Load the file `nscg17small.dta`. You can find the data in the `Data` folder. 
   + This data is a subset of the National Survey of College Graduates (NSCG) 2017, which collects data on the educational and occupational characteristics of college graduates in the United States.

2. Each row corresponds to a unique respondent. Let's create a new column called "ID". There are various ways to create an ID column. Here, let's create an ID column that starts from 1 and increments by 1 for each row.
3. To take a quick look at the summary statistics of a specific column, `summary()` function is useful. Use `summary()` to create a table of the descriptive statistics for salary. You'll provide salary column to `summary()` as a vector. 
4. Create a new variable in your data that represents the z-score
of the hours worked (use `hrswk` variable).
$$Z = (x - \mu)/\sigma$$
, where $Z = \text{standard score}$, $x =\text{observed value}$, $\mu = \text{mean of sample}$, and $\sigma = \text{standard deviation of the sample}$.
5. Calculate the share of observations in your data sample with
above average hours worked.


## {{< fa lightbulb >}} Solutions

```{r}
# === Part 1 === #
library(rio)
nscg17 <- import("Data/nscg17small.dta")

# === Part 2 === #
nscg17$ID <- 1:nrow(nscg17)

# === Part 3 === #
summary(nscg17$salary)

# === Part 4 === #
nscg17$z_hrswk <- (nscg17$hrswk - mean(nscg17$hrswk)) / sd(nscg17$hrswk)
# or using with() function, you can write the code more concisely
# nscg17$z_hrswk2 <- with(nscg17, (hrswk - mean(hrswk)) / sd(hrswk))

# Note: For part 2 and 3, you can use within() function to create new columns more concisely.
# nscg17 <- 
#   within(
#     nscg17, {
#       ID <- 1:nrow(nscg17)
#       z_hrswk <- (hrswk - mean(hrswk)) / sd(hrswk)
#     }

# === Part 5 === #
# create a logical vector that indicates whether the hours worked is above average
above_avg_hrswk <- with(nscg17, z_hrswk > mean(z_hrswk)) # you can get the same result by using `hrswk`.
# subset the data
nscg17_above_avg_hrswk <- nscg17[above_avg_hrswk, ]
# calculate the share of observations with above average hours worked
share_above_avg_hrswk <- nrow(nscg17_above_avg_hrswk) / nrow(nscg17)
share_above_avg_hrswk
```

<!-- end panel: ex  --> 
:::

<br>


<!-- # Exercise (Advanced) -->

<!-- start panel: ex --> 
<!-- :::{.panel-tabset .nav-pills}  
## {{< fa person-chalkboard >}} Instructions  

Let's use `flight.rds` data in the `Data` folder. Load the data. We will only look at the flight data with "MSP" (Minneapolis–Saint Paul International Airport) as the destination airport. 

**Question**

1. Create `flights_mn` data that contains the column `year`, `month`, `day`, `dep_delay`, `arr_delay`, `dest`, `carrier` and subset rows that have "MSP" as `dest`. 

2. Using `flights_mn`, find the proportion of flights that are delayed more than 30 minutes in total for each each airline company (`carrier`).

::: {.callout-note}
+ You can solve this exercise problem by using only base-R functions. See Hints if you need help.
+ Also, note that googling is one of the most important skills for coding!
:::

## {{< fa message >}} Hints (step by step)

Here are the steps that I took to solve the problem. You don't need to exactly follow these steps.

1. Load the `flight.rds` data in the `Data` folder.
2. To create `flights_mn`, select the columns `year`, `month`, `day`, `dep_delay`, `arr_delay`, `dest`, and  `carrier`, and subset rows that have "MSP" as `dest`.



## {{< fa person-chalkboard >}} Solutions

```{r}
#| eval: false
# === Step 1: Load data === #
flights <- readRDS("Data/flight.rds")

# === Part 2: subset data === #
# define the columns to select
select_cols <- c("year", "month", "day", "dep_delay", "arr_delay", "dest", "carrier")
flights_mn <- flights[flights$dest == "MSP", select_cols]
# or
# flights_mn <- with(flights, flights[dest == "MSP", select_cols])

flights_mn$total_delay <- flights_mn$dep_delay + flights_mn$arr_delay

delayed <- flights_mn$total_delay > 30

total_flights <- table(flights_mn$carrier)

delayed_flights <- table(flights_mn$carrier[delayed], row.names = unique(flights_mn$carrier))

# Calculate the proportion of delayed flights
proportions <- delayed_flights / total_flights

# Print the result
print(proportions)
``` -->

<!-- end panel: ex  --> 
<!-- ::: -->